---
title: "Final_Project"
output: html_document
date: "2025-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Evaluation of CDC45 as a Prognostic Biomarker in Cervical Cancer

The following analysis aims to reproduce some of the results presented in the study "Expression and Prognosis of CDC45 in Cervical Cancer Based on the GEO Database" by Liu et al. (2021), which evaluated the expression and prognostic value of the CDC45 gene in cervical cancer. For this purpose, the gene expression dataset available in the Gene Expression Omnibus (GEO) repository, under accession number GSE63514, was used.
4
# PART 1: 
In this section the code shared on the following site was used: https://www.ncbi.nlm.nih.gov/geo/geo2r/?acc=GSE63514

```{r,echo=FALSE}
library(GEOquery)
library(limma)
library(umap)
library(ggplot2)
library(reshape2)
```


```{r}
# Loading dataset GSE63514 
#(GSEMatrix = TRUE, indicates that we are downloading the gene expression matrix)

gset <- getGEO("GSE63514", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL570", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

#express: Extracts the gene expression matrix from the gset object, creates a new object called ex
ex <- exprs(gset)
dim(ex)
#rows 54675   columns:128
# see if we have NAs:
sum(is.na(ex))
#0 na

#A logarithmic transformation (log2) is performed
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
ex <- log2(ex) }

# box-and-whisker plot
dev.new(width=3+ncol(gset)/6, height=5)
par(mar=c(7,4,2,1))
title <- paste ("GSE63514", "/", annotation(gset), sep ="")
boxplot(ex, boxwex=0.7, notch=T, main=title, outline=FALSE, las=2)
dev.off()

# expression value distribution plot
par(mar=c(4,4,2,1))
title <- paste ("GSE63514", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, main=title, legend=F)

# mean-variance trend
ex <- na.omit(ex) # eliminate rows with NAs
plotSA(lmFit(ex), main="Mean variance trend, GSE63514")

```

## PART 2: 


```{r}
# Perform Differential Expression Analysis with limma

# Create a design matrix based on the condition labels
design <- model.matrix(~ condition, data = pData(gset))

# Fit a linear model
fit <- lmFit(ex, design)

# Apply empirical Bayes moderation to the fit model
fit2 <- eBayes(fit)

# Step 6: Extract results
# We are interested in comparing the 'Cancer' condition to 'Normal'
results <- topTable(fit2, adjust = "fdr", number = Inf)

# Print the top results to check if they make sense
head(results)

# Step 7: Filter results for significantly differentially expressed genes (e.g., FDR < 0.05)
significant_genes <- results[results$adj.P.Val < 0.05, ]
```


## PART 3: 
Exploratory data analysis and data visualization:
The characteristics of the dataset samples are explored to identify and classify each one as "Normal", "Cancer" or "Precancer" according to its description.


```{r}
# Verify first rows:
head(pData(gset)$characteristics_ch1)

# Filter 
cancer_samples <- grep("cancer", pData(gset)$characteristics_ch1, ignore.case = TRUE)
cancer_samples

#Displays all unique descriptions that appear in the characteristics_ch1 column of the gset object
unique(pData(gset)$characteristics_ch1)

# Assign condition labels based on different characteristics
condition_labels <- ifelse(grepl("normal", pData(gset)$characteristics_ch1, ignore.case = TRUE), "Normal",
                           ifelse(grepl("cancer", pData(gset)$characteristics_ch1, ignore.case = TRUE), "Cancer", "Precancer"))

# Assign the above labels to the gset object
pData(gset)$condition <- condition_labels

# Print the results in a table
table(pData(gset)$condition)

```
# Data visualization:
1- Barplot/ Lollipop chart
2- Boxplot/ Violin plot

```{r}
# Count the number of samples in each condition
condition_counts <- table(pData(gset)$condition)

# Convert the table to a data frame and rename the columns
condition_counts_df <- as.data.frame(condition_counts)
colnames(condition_counts_df) <- c("Condition", "Frequency")

# Inspect the first few rows to verify column structure
head(condition_counts_df)

# Generate a bar plot showing the number of samples per condition

library(viridis)  #'viridis' palette for colorblind-friendly
ggplot(condition_counts_df, aes(x = Condition, y = Frequency, fill = Condition)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Sample Distribution by Condition", x = "Condition", y = "Number of Samples") + 
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal()

# Alternative visualization: lollipop chart
ggplot(condition_counts_df, aes(x = Condition, y = Frequency, color = Condition)) + 
  geom_segment(aes(x = Condition, xend = Condition, y = 0, yend = Frequency), linewidth = 1) + 
  geom_point(size = 3) + 
  labs(title = "Sample Distribution by Condition (Lollipop Chart)", 
       x = "Condition", y = "Sample Count") + 
  theme_minimal() + 
  scale_color_manual(values = c("Normal" = "blue", "Precancer" = "orange", "Cancer" = "red"))

# Prepare expression data in long format for plotting
expression_data_long <- data.frame(
  Gene = rep(rownames(ex), each = ncol(ex)),
  Expression = as.vector(t(ex)),
  Condition = rep(pData(gset)$condition, times = nrow(ex))
)

# Boxplot: visualize gene expression distribution across conditions
ggplot(expression_data_long, aes(x = Condition, y = Expression, fill = Condition)) +
  geom_boxplot() +
  labs(title = "Gene Expression Distribution by Condition (Boxplot)", 
       x = "Condition", y = "Gene Expression") +
  theme_minimal()

# Violin plot: alternative visualization for gene expression distribution
ggplot(expression_data_long, aes(x = Condition, y = Expression, fill = Condition)) +
  geom_violin() +
  labs(title = "Gene Expression Distribution by Condition (Violin Plot)", 
       x = "Condition", y = "Gene Expression") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")

```

# PART 4
PCA Analysis: Normal vs Cancer
Principal components (PCs) are used to identify variable dependencies, explore relationships among individuals, stabilize estimates, assess multivariate normality, and detect outliers.
This technique is widely used in Transcriptomics and results very helpful for exploring complex gene expression data and revealing underlying patterns.

```{r}

# Filter samples for "Cancer" and "Normal"
selected_samples <- pData(gset)$condition %in% c("Cancer", "Normal")

# Extract and transpose the expression matrix only with the selected samples
exprs_filtered <- exprs(gset)[, selected_samples]

# Perform PCA
pca <- prcomp(t(exprs_filtered), scale. = TRUE)

# Create a dataframe with the PCA scores
#PCA scores are necessary to calculate, as they represent the projections of the samples onto the principal components. They are used to create PCA plots by positioning the samples in the reduced space based on their scores along the first principal components (PC1 and PC2).
pca_scores <- data.frame(pca$x)

# Add the conditions (only Cancer and Normal)
pca_scores$condition <- pData(gset)$condition[selected_samples]

# Calculate the variance explained by each component
explained_var <- summary(pca)$importance[2, 1:2] * 100  
explained_var


#PCA PLOT
ggplot(pca_scores, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 2) +
  labs(
    title = "PCA Plot of GSE63514 (Cancer vs Normal)",
    x = paste0("PC1 (", round(explained_var[1], 2), "%)"),
    y = paste0("PC2 (", round(explained_var[2], 2), "%)")
  ) +
  theme_minimal()

```


References:
Liu, J., Li, Y., Liu, Q., & Li, Y. (2021). Expression and prognosis of CDC45 in cervical cancer based on the GEO database. PeerJ, 9, e12114. https://doi.org/10.7717/peerj.12114




#PART 5:
Volcano plot and Heatmap

```{r}

#Volcano plot
#To build the volcano plot graph it is necessary to classify the results: significant (pv<0.05), and logFC>1, otherwise for non-significant ones.
gset <- getGEO("GSE63514", GSEMatrix = TRUE, getGPL = FALSE)
if (length(gset) > 1) idx <- grep("GPL570", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# Extraer la matriz de expresión
ex <- exprs(gset)
# Assign conditions (Normal vs Cancer)
condition_labels <- ifelse(grepl("normal", pData(gset)$characteristics_ch1, ignore.case = TRUE), "Normal",
                           ifelse(grepl("cancer", pData(gset)$characteristics_ch1, ignore.case = TRUE), "Cancer", "Other"))
pData(gset)$condition <- condition_labels  # Assign condition labels to pData

# Check that the 'condition' column is correctly created
head(pData(gset)$condition)

# Create the design matrix for differential analysis
design <- model.matrix(~ 0 + condition, data = pData(gset))  # Use pData(gset) to access 'condition'
colnames(design) <- levels(factor(pData(gset)$condition))

# Proceed with differential analysis
fit <- lmFit(ex, design)

# Perform the contrast between Cancer and Normal
contrast.matrix <- makeContrasts(Cancer - Normal, levels = design)

# Adjust the contrasts and perform the analysis
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# Get the results for the Cancer vs Normal contrast
results <- topTable(fit2, coef = 1, n = Inf)  # coef=1 for Cancer - Normal

# Create the "significance" column for the Volcano Plot
results$significance <- ifelse(results$P.Value < 0.05 & abs(results$logFC) > 1, "Significant", "Not Significant")

# Generate the Volcano Plot
ggplot(results, aes(x = logFC, y = -log10(P.Value), color = significance)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(values = c("red", "gray")) +
  labs(
    title = "Volcano Plot: Cancer vs Normal",
    x = "Log2 Fold Change (Cancer vs Normal)",
    y = "-Log10(p-value)"
  ) +
  theme_minimal() +
  theme(legend.position = "top")






```

```{r}

#Heatmap
library(pheatmap)

top_genes <- head(order(results$P.Value), 30)
exprs_top_genes <- exprs_filtered[top_genes, ]

dim(exprs_top_genes)

length(condition)

unique(condition)

sum(is.na(condition))

condition <- factor(condition, levels = c("Normal", "Cancer"))

levels(condition)

annotation_data <- data.frame(Condition = condition)

head(annotation_data)

dim(exprs_top_genes)

dim(annotation_data)

annotation_colors <- list(Condition = c("Normal" = "blue", "Cancer" = "red"))

pheatmap(exprs_top_genes, 
         scale = "row",  
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean",  
         clustering_method = "complete",  
         show_rownames = FALSE,  
         show_colnames = TRUE,  
         annotation_col = annotation_data,  
         annotation_colors = annotation_colors)

```




References:
Liu, J., Li, Y., Liu, Q., & Li, Y. (2021). Expression and prognosis of CDC45 in cervical cancer based on the GEO database. PeerJ, 9, e12114. https://doi.org/10.7717/peerj.12114