---
title: "Data_exploration"
output: html_document
date: "2025-04-16"
---

Data exploration of the data set GSE63514 

```{r}
# Loading dataset GSE63514 
#(GSEMatrix = TRUE, indicates that we are downloading the gene expression matrix)
library(GEOquery)
library(limma)
library(umap)
library(ggplot2)
library(RColorBrewer)
library(reshape2)

gset <- getGEO("GSE63514", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL570", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]
```


```{r}
# See the first and last rowns
head(exprs(gset),2)
tail(exprs(gset),2)

#Rowns and colums number
dim(exprs(gset))

# Column's names
colnames(exprs(gset))
#The colnames start with GSM (corresponds to samples)


```
In the dataset, the rows represent probeset IDs from a microarray (e.g., 1007_s_at), while the columns represent samples identified by "GSM" (Gene Expression Sample). Each value indicates the gene expression level for a specific probe in a given sample.


Histogram of gene expression 
```{r}

hist(exprs(gset), main="Gene Expression Distribution ", xlab="Expression Value", col="pink", breaks=50)
```


The histogram suggests a left-skewed distribution. This is quite common in gene expression data. That is, in this case, most genes have low expression levels, indicating that only a small subset has high expression.


Boxplot

```{r}
boxplot(exprs(gset), main="Gene Expression", xlab="Samples", ylab="Expression Value", col="green")
```

